---
title: "Finance Quantitative"
subtitle: "Calcul Actuariel"
author: Patrick Hénaff
date: "Version: `r format(Sys.Date(), '%d %b %Y')`"
output:
  pdf_document:
    keep_tex: false
    fig_caption: yes
    latex_engine: pdflatex
geometry: margin=1in

header-includes:
  - \usepackage[utf8]{inputenc}
  - \usepackage{eurosym}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
```

```{r load-libraries, include=FALSE, echo=TRUE}
library(quantmod)
library(xts)
library(hornpa)
library(lubridate)
library(xtable)
library(PerformanceAnalytics)
library(TTR)
library(SuppDists)
library(lubridate)
library(roll)
library(Hmisc)
library(nFactors)
library(kableExtra)
library(quadprog)
```

# Time Value of Money (Van Horne)
 
Consider the following cash flow streams:

\begin{table}[ht]
  \centering
\begin{tabular}{llllll}
\toprule \\
Year & 1 & 2 & 3 & 4 & 5 \\
\midrule
W & 100 & 200 & 200 & 300 & 300 \\
X & 600 &     &     &     & \\
Y &     &     &     &     & 1200 \\
Z & 200 &     & 500 &     & 300 \\
\bottomrule
\end{tabular}
\end{table}

\begin{enumerate}
  \item Compute the future value of each stream at a compounded rate of 10\%.
  \item Compute the present value of each stream at a compounded rate of 14\%.
\end{enumerate}

```{r}
df = data.frame(
  "W" = c(100, 200, 200, 300, 300),
  "X" = c(600, 0, 0, 0, 0),
  "Y" = c(0, 0, 0, 0, 1200),
  "Z" = c(200, 0, 500, 0, 300)
)

r_fut = 0.10
r_pres = 0.14

func_futval <- function(col) {
  sum(col * (1 + r_fut)^(rev(0:length(col)-1)))
}

func_presval <- function(col) {
  sum(col / (1 + r_pres)^(1:length(col)))
}

```

```{r, warning=FALSE}
fut_df = apply(df, MARGIN=2, FUN=func_futval)
fut_df
```

```{r}
pres_df = apply(df, MARGIN=2, FUN=func_presval)
pres_df
```


# Comparing contracts (Van Horne)

On a contract, you have a choice of receiving 25,000 \euro in six years or 50,000 \euro in 12 years. At which compound annual rate would you be indifferent between the two options?

$\frac{FV_{12Y}}{(1+r)^6} = FV_{6Y} \Leftrightarrow r = \left( \frac{FV_{12Y}}{FV_{6Y}} \right)^{\frac{1}{6}} - 1$

```{r}
FV_12Y = 50000
FV_6Y = 25000

r_eq=  (FV_12Y / FV_6Y)^(1/6) - 1
r_eq
```

# Mortgage (Van Horne)

You obtain a 10-year, 50,000 \euro\  loan. The compound annual interest rate is 8\%. The loan is paid back by 10 annual installments of 7,451.47 \euro. 

\begin{enumerate}
\item How much of the first year payment is principal?
\item How much total interest will be paid over the life of the loan?
\end{enumerate}

```{r}
FV = 50000
r = 0.08
nb_years = 10
annuity = 7451.74

principal_1 = annuity - FV*r
principal_1
```
```{r}
total_int = annuity * nb_years - FV
total_int
```

# Savings Plan

You need to have 50,000 \euro\  at the end of ten years. To accumulate this sum, you plan to save a certain amount at the end of each year, for the next ten years. The bank pays 8\% interest, compounded annually. How much should you save each year?

$P \sum_{i=1}^{n} (1+r)^i = FV \Leftrightarrow P \frac{(1+r)^n-1}{r}=FV \Leftrightarrow P = FV \frac{r}{(1+r)^n-1}$

```{r}
P = FV * r / ((1 + r)^nb_years - 1)
P
```

# Mortgage Paydown

You have borrowed 14,300 \euro\  at a compound annual interest rate of 15\%. You can make annual payments of 3,000 \euro\  on your loan. How long will it be before your loan is completely paid down?

$P \sum_{i=1}^{n} \frac{1}{\left(1+r\right)^i} = FV \Leftrightarrow P \frac{1+r}{r} \left(1 - \frac{1}{\left(1+r\right)^n}\right) = FV \Leftrightarrow n = \frac{\ln\left(1 - \frac{FV r}{P \left(1 + r\right)}\right)}{\ln\left(\frac{1}{1 + r}\right)}$

```{r}
FV = 14300
r = 0.15
P = 3000

n = log(1 - FV * r / (P * (1 + r))) / log(1 / (1 + r))
n
```


